\documentclass{amsart}

\usepackage{amsmath}
\usepackage{amssymb}

\begin{document}

\newcommand\meet{\mathrm{meet}}

The meet computation in typclingo.

Let $V$ be a set of \emph{type variables},
$F$ be a set of \emph{function constructors}, and
$T$ be a set of \emph{type constants}.

The set $\mathcal{T}$ of \emph{type expressions} is the smallest set containing:
\begin{itemize}
	\item $\top$ (the top type)
	\item $e$ for $e \in T \cup V$
	\item $f(e_1,\dots,e_n)$ for $e_1,\dots,e_n \in \mathcal{T}$, $f \in F$, and $n \geq 0$ (function types)
	\item $(e_1 \mid ... \mid e_n)$ for $e_1,\dots,e_n \in \mathcal{T}$ and $n \geq 0$ (type union)
\end{itemize}

We omit the parentheses for function types if the arity is zero
and type unions when there is no risk of ambiguity.
We define $\bot$ as the empty type union.

A \emph{type specification} $S$ is a set of expressions $t \lessdot e$ or $t \doteq e$
such that $t \in T$ and $e \in \mathcal{T}$ where $e$ does not contain type variables.

We assume that the type constants always contain $\mathbb{F}$, $\mathbb{N}$ and $\mathbb{S}$.
Furthermore, type specifications always include:
\begin{itemize}
	\item $\mathbb{F} \lessdot \top$ (the type containing all function types)
	\item $\mathbb{N} \lessdot \top$ (the type containing all numbers)
	\item $\mathbb{S} \doteq \mathbb{F} \mid \mathbb{N}$
\end{itemize}

Type specifications may define recursive types, but every (recursive) type must have a base case.
Each type constant occurring on the right-hand side of a $\lessdot$ or $\doteq$ is defined exactly once.


We let $\Sigma$ be the set of all \emph{type variable mappings} from type variables to type expressions.
We assume that such a mapping $\sigma$ maps all variables in $V$ to some type expression.
We denote a type mapping as a set of expressions of the form $v \mapsto e$ where $v \in V$ and $e \in \mathcal{T}$.
A mapping maps type variables to $\mathbb{S}$ if not explicitly given.
Given $\sigma_1, \sigma_2 \in \Sigma$, we let $\sigma_1 \sqcup \sigma_2$ be the mapping defined as:
$(\sigma_1 \sqcup \sigma_2)(v) = \sigma_1(v) \mid \sigma_2(v)$ for all $v \in V$.

Function $\mathrm{meet}_{S}: \mathcal{T} \times \mathcal{T} \times \Sigma \times C \mapsto \mathcal{T} \times \Sigma$
computes the \emph{meet} of two type expressions given a type variable mapping, a type specification $S$, and a seen set $C$.
The result is a new type expression and an updated type variable mapping.
Here $C \subseteq \mathcal{T} \times \mathcal{T}$ is a set of pairs of type expressions for which the meet has already been computed.

We define $\mathrm{meet}_{S}$ as follows:
\begin{itemize}
    \item $\meet_S(e_1, e_2, \sigma, C) = (\bot, \sigma)$ where $(e_1, e_2) \in C$ or $(e_2, e_1) \in C$
    \item $\meet_S(e,e,\sigma,C) = \meet_{S}(e, \top, \sigma, C) = \meet_{S}(\top, e, \sigma, C) = (e, \sigma)$
    \item $\meet_S(t, e_1, \sigma, C) = \meet_S(e_1, t, \sigma, C) = \meet_S(e_1,e_2,\sigma,C')$\\
        where $t \doteq e_2 \in S$ and $C' = C \cup \{(t, e_1)\}$
    \item $\meet_{S}(e_1, e_2, \sigma, C) = \meet_{S}(e_2, e_1, \sigma, C) = (e', \{v' \mapsto \sigma'(v') \mid v' \neq v \in V\} \cup \{v \mapsto e'\})$\\
        where $v \in V$, $C' = C \cup \{(e_1, e_2)\}$, and $(e', \sigma') = \meet_{S}(\sigma(v), e_2, \sigma, C')$
    \item $\meet_S(e, e_1 \mid \dots \mid e_n, \sigma, C) = \meet_S(e_1 \mid \dots \mid e_n, e, \sigma, C) = (e_{i_1}' \mid \dots \mid e_{i_k}', \sigma_{i_1} \sqcup \dots \sqcup \sigma_{i_k})$ \\
        where $C' = C \cup \{(e, e_1 \mid \dots \mid e_n)\}$, $(e_i',\sigma_i) = \meet_S(e_i,e,\sigma,C')$, and $i_j$ are all the indices such that $e_{i_j}' \neq \bot$
    \item $\meet_S(f(e_1,\dots,e_n),\mathbb{F},\sigma,C) = (f(e_1,\dots,e_n), \sigma)$
    \item $\meet_S(f(e_1,\dots,e_n),t,\sigma,C) = \begin{cases} (t,\sigma') & e' \neq \bot \\ (\bot,\sigma') & \text{otherwise} \end{cases}$\\
        where $t \lessdot e \in S$, $t \neq \mathbb{N}$, $t \neq \mathbb{F}$, $C' = C \cup \{(f(e_1,\dots,e_n), t)\}$, and $(e',\sigma') = \meet_S(f(e_1,\dots,e_n),e,\sigma,C')$
    \item $\meet_S(f(e_1,\dots,e_n),f(g_1,\dots,g_n), \sigma, C) = \begin{cases} (f(h_1,\dots,h_n),\sigma_n) & \text{all } h_i \neq \bot \\ (\bot,\sigma_n) & \text{otherwise} \end{cases}$
        where $C' = C \cup \{(f(e_1,\dots,e_n), f(g_1,\dots,g_n))\}$, $\sigma_0 = \sigma$ and $(h_i, \sigma_i) = \meet_S(e_i,g_i,\sigma_{i-1}, C')$
    \item $\meet_S(t_1,t_2,\sigma,C) = \begin{cases} (t_2, \sigma) & t_1 \text{ is a reachable super type of } t_2 \text{ in } S \\ (t_1, \sigma) & t_2 \text{ is a reachable super type of } t_1 \text{ in } S \\ (\bot,\sigma) & \text{otherwise} \end{cases}$\\
        where $t_1 \lessdot e_1, t_2 \lessdot e_2 \in S$ and $C' = C \cup \{(t_1, t_2)\}$
    \item $\meet_S(e_1,e_2,\sigma,C) = (\bot,\sigma)$, otherwise
\end{itemize}

Note that the seen set prevents infinite recursion when computing meets of recursive types.
Each pair of type expressions is seen at most once, so the computation always terminates.

To give an idea how the meet computation is applied in typclingo,
let $p(X), q(X)$ be some rule body.
We assume that we have some type specification $S$ containing type definitions $e_1$ and $e_2$
for the arguments of $p$ and $q$, respectively.
We can then map the atoms $p(X)$ and $q(X)$ to the type expressions $p(x)$ and $q(x)$ where $x$ is a type variable.

To check that the body is well-typed,
we compute $\mathrm{meet}_{S}(p(x), p(e_1),\emptyset,\emptyset) = (e,\sigma)$.
In the next step, we compute $\mathrm{meet}_{S}(q(x), q(e_2),\sigma,\emptyset) = (e',\sigma')$.
If both $e$ and $e'$ are not $\bot$,
then the body is well-typed and the type of $X$ is given by $\sigma'(x)$.
\end{document}
